---
title: "Google Data Analytics Certificate Capstone (Bellabeat Case Study)"
author: "Gabriel Fernandez"
output:
  
  pdf_document:
  github_document:
urlcolor: blue
---



```{r}
# Set default options for code chunks
knitr::opts_chunk$set(
  echo = FALSE,
  include = FALSE,# Do not display R code and its output
  comment = NA,          # Suppress code comments in output
  warning = FALSE,       # Suppress warning messages
  fig.align = 'center',  # Align figures in the center
  eval = TRUE            # Evaluate R code
)

```


```{r warning=FALSE, message=FALSE}
# Import libraries 
library(tidyverse) # includes ggplot2
library(skimr) #  provides a compact and informative summary of your dataframe or dataset
library(lubridate)
library(janitor) # set of utility functions for data cleaning and data frame tidying tasks
library(RColorBrewer) # Color palettes for data visualization
library(ggcorrplot) # Visualize correlation matrices using ggplot2
library(scales) # formatting and transforming data for visualizations

```


Let us clean:

- Change column names to lower case because R is case sensitive.

- Change "Id" from double to a character because the number represents a category.

- Change "ActivityDate" from char to date.


# heading 1
```{r }
weight_logs <-
  read_csv("original_data/weightLogInfo_merged.csv",
    trim_ws = TRUE,
    show_col_types = FALSE
  )
```

## Heading 2

```{r}
glimpse(weight_logs)
```



```{r }
glimpse(weight_logs)
```



# References

### Guidelines and research articles

- 1. Handling sedentary time: [A Comparison of Sedentary Behavior as Measured by the Fitbit and ActivPAL in College Students](https://www.mdpi.com/1660-4601/18/8/3914)

- 2. Danger of prolong sitting(sedentary time): [Association of daily sitting time and leisure-time physical activity with body fat among U.S. adults. Journal of Sport and Health Science](https://www.sciencedirect.com/science/article/pii/S2095254622001016)

- 3. [Dietary Guidelines for Americans, 2020-2025](https://www.dietaryguidelines.gov/)

- 4. [Physical Activity Guidelines for Americans (2nd ed.)](https://health.gov/sites/default/files/2019-09/Physical_Activity_Guidelines_2nd_edition.pdf)


### Links
- [Projects Datasets:](https://www.kaggle.com/datasets/arashnic/fitbit)

- [EDA guide](https://rpubs.com/jovial/r) 

- Metadata: [Fitbit data dictionary](https://www.fitabase.com/resources/knowledge-base/exporting-data/data-dictionaries/)

- [Plotting histograms with ggplot2](https://appsilon.com/ggplot2-histograms/)

- [Histograms article](https://statisticsbyjim.com/basics/histograms/)
- [Error bars vs CI](https://blogs.sas.com/content/iml/2019/10/09/statistic-error-bars-mean.html)


- [Add density line to histogram](https://r-coder.com/density-plot-r)

- [Categorical, ordinal, interval,  variables](https://www.graphpad.com/guides/prism/latest/statistics/the_different_kinds_of_variabl.htm)


### Appendix: Interesting sites for further investigation

- [Adult Physical Inactivity Prevalence Maps by Race/Ethnicity](https://www.cdc.gov/physicalactivity/data/inactivity-prevalence-maps/index.html#Race-Ethnicity)

- [Physical activity among adults aged 18 and over : United States, 2020](https://stacks.cdc.gov/view/cdc/120213)


"Sleep time is not considered sedentary time, so it was removed to determine the waking day and to allow the proportion of the day spent sedentary to be calculated." [[Reference 1]](https://www.mdpi.com/1660-4601/18/8/3914)













